The National Eutrophication Survey: lake characteristics and historical nutrient concentrations

### 2017-11-07 08_manuscript/template.tex
add bibtex

%% 2-column papers and discussion papers
\documentclass[journal abbreviation, manuscript]{copernicus}

@@

\introduction  %% \introduction[modified heading if necessary]
Effective management of inland freshwater lakes requires an understanding of the factors that affect water quality and how these factors change over time. One of these factors, termed eutrophication, occurs when excess nutrient inputs from human activities fuels increases in algal growth which can cause hypoxia and decreases in water clarity. Eutrophication of surface waters from increased phosphorus and nitrogen loading has been observed in connection with altered land-use especially in areas of rapid urbanization and intensive agriculture (Smith et al., 1999, Smith et al., 2014). As human populations and their impacts continue to grow, eutrophication is expected to become more widespread (Bennett et al., 2001, Taranu and Gregory-Eaves, 2008). Historical datasets are needed in order to track, understand, and manage eutrophication in lakes and reservoirs because they serve as an important baseline for modern studies.

The U.S. Environmental Protection Agency (EPA) designed and implemented the National Eutrophication Survey (NES) in order to investigate the extent of eutrophication in freshwater lakes and reservoirs across the contiguous United States (US). Sampling took place in over 800 lakes and reservoirs from 1972 to 1975, and included a variety of physical, chemical, and biological metrics including data on nutrients and nutrient loading, hydrologic retention time, morphometry, and plankton community diversity.  Unlike current EPA National Lake Assessments that select a random sample of lakes across the US, the NES  targeted only lakes impacted directly or indirectly by municipal sewage treatment plant discharge (USEPA 2009). Until recently, these data were only available in their entirety as four separate scanned reports representing the northeastern and northcentral  (northeastern), eastern and southeastern (southeastern), central, and western regions of the US (Figure 1). In the remainder of the present paper we refer to the former two regions as simply the northeastern and southeastern regions.

@@

To our knowledge, there have been no attempts to transcribe the data into a usable, searchable digital database despite its use in previous studies.   For example, large portions of the dataset were used to examine large scale relationships between residence time and phytoplankton abundance (Soballe and Kimmel, 1987). Also, it was used to predict eutrophication incidence in a Bayesian framework (Lamon and Stow 2004). Smaller portions of the data were used to explore drivers of nutrient loading (Stomp et al., 2011, Brett and Benjamin, 2007). Yet, to our knowledge, the only study to use the NES dataset and provide a publicly available data supplement is Stomp et al., (2011), and their data supplement was limited to a small subset of the available variables relating to phytoplankton community diversity.

@@

Data was collected from multiple locations within the water column and included in-situ measurements as well as laboratory analysis. Flow estimates and drainage area calculations were provided by the USGS and were determined from flow gages when present.  More detailed information on sampling methods, units, equipment, and accuracy can be found in the EPA survey methods publication (USEPA, 1975). Due to historical nature of the dataset, the NES sampling design differs from more modern efforts (USEPA, 2009). For example, the original NES data was collected from four separate regions of the US over the course of four years, whereas current assessments complete nation-wide sampling in a single summer. As such, NES data values represent the median of measurements taken in the spring, summer, and fall in either 1972 (northeastern), 1973 (southeastern), 1974 (central), or 1975 (western) rather than measurements taken in a single year.

We obtained the NES archival scanned reports from the EPA National Service Center for Environmental Publications (available at: https://www.epa.gov/nscep). The data for each NES region is contained in four separate files. We extracted the data from each file using automated techniques followed by manual quality assurance and checking. To begin, we enhanced (de-noised) each file using the local adaptive filtering algorithm as provided by the Imagemagick program (v6.8.9-9, available at: https://www.imagemagick.org/). Next, we processed the enhanced files using the Tesseract optical character recognition program (OCR) (Ooms, 2017, Smith, 2007). The output of these initial extraction steps were recorded in a set of “raw data” files where each file contains the raw unprocessed text of each document page. The contents of specific fields in the raw data were extracted to a database using the automated rules provided by the nesR software package (Stachelek, 2017a). Finally, all values in the database were manually checked for accuracy against the original scanned reports. Inaccurate OCR outputs were hand-corrected in the final database.

We provide the final dataset in an open non-proprietary format (comma-delimited, *.csv).  We generated metadata descriptions from the contents of the original scanned reports. All calculations, table construction, and figure generation were performed in R and saved as reproducible R scripts (R Core Team, 2017). Table and figure generation was accomplished with the use of the reshape2, plyr, and sp packages (Wickham, 2007, Wickham, 2011, Pebesma and Bivand, 2005).

@@

We have demonstrated an approach for rescuing historical data from scanned documents. In particular, our approach involved a two-step process of automated data scraping followed by hand-curation and quality assurance. Overall, we found that optical character recognition was an efficient method for reducing the labor associated with transcribing analog text records (Drinkwater et al., 2014). Unfortunately, optical character recognition technology does not have absolute accuracy. In our case, transcription was hampered by poor print and scan quality of the source paper documents. We discovered through our manual validation procedure that the OCR computations produced inaccurate values in approximately 6 - 17\% of the cells in the complete dataset (n = 4836). We expect that accuracy could be improved by experimenting with varying the window size of the local adaptive thresholding algorithm relative to the document font size. Our ability to experiment with thresholding window size was limited due to the computationally expensive nature of these extractions.

The end result of our approach was data from every lake and nearly every variable in the NES survey dataset. The only primary subset of the NES data which is not included in our final product is the phytoplankton distribution data which has already been digitally transcribed by Stomp et al. (2011). The results of the present study could be used to explore anthropogenic and environmental drivers of lake eutrophication as well as to verify previously documented trends. One example is the 2007 National Lakes Assessment (NLA) Report, which included a reanalysis of some of the NES study lakes (USEPA, 2009). This reanalysis considered population level trends in the NES lakes but did not consider trends in individual lakes or potential environmental drivers contributing to observed trends. On a population basis, the NLA reanalysis found that less than 30\% of the NES lakes had increased chlorophyll and phosphorus concentrations. The results of the present study could be used to verify these claims as well as to compare the NES data with more recent work such as the 2012 National Lakes Assessment.  Note that sampling techniques may differ from current techniques, so care should be given when making comparisons. In addition to their utility in validating historical trends, this dataset has value because it contains data on a number of hydrographic variables which are difficult to estimate such as water residence (retention) time. Such data is critical to a variety of hydrological and water quality modelling efforts (Brett and Benjamin, 2008).

Although our goal was to digitally transcribe the full NES dataset to facilitate studies on historical nutrient loading, it is worth noting the similarities between the present study and other scientific record digitization initiatives. Such initiatives are common in the climate and ocean sciences but they are just starting to gain momentum in the biological sciences (Allan et al., 2011, Freeman et al., 2016). To our knowledge, the present study is the first large-scale attempt at digitization of historical limnology records. We hope that by making our analysis open and reproducible we will inspire future efforts at recovering important records from the pre-digital era.

@@

\codedataavailability{
Original scanned reports from the EPA are available from the EPA National Service Center for Environmental Publications (https://www.epa.gov/nscep).  Our cleaned and useable data are available for download at doi:10.5063/F10G3H3Z. The data are provided as a zip file which contains all versions of the data including the raw and quality checked versions (Stachelek et al., 2017b). Moreover, the R package and R code used to scrape and analyze the data are provided by Stachelek (2017a) so that the methods may be reproduced and openly available for (re)use. All figures and summary statistics were generated by R scripts available in the data supplement linked above.
} %% use this section when having data sets and software code available

>>>>

%% 2-column papers and discussion papers
\documentclass[journal abbreviations, manuscript]{copernicus}

@@

\introduction  %% \introduction[modified heading if necessary]
Effective management of inland freshwater lakes requires an understanding of the factors that affect water quality and how these factors change over time. One of these factors, termed eutrophication, occurs when excess nutrient inputs from human activities fuels increases in algal growth which can cause hypoxia and decreases in water clarity. Eutrophication of surface waters from increased phosphorus and nitrogen loading has been observed in connection with altered land-use especially in areas of rapid urbanization and intensive agriculture \citep{smith1999eutrophication, smith2014comment}. As human populations and their impacts continue to grow, eutrophication is expected to become more widespread \citep{bennett2001human, taranu2008quantifying}. Historical datasets are needed in order to track, understand, and manage eutrophication in lakes and reservoirs because they serve as an important baseline for modern studies.

The U.S. Environmental Protection Agency (EPA) designed and implemented the National Eutrophication Survey (NES) in order to investigate the extent of eutrophication in freshwater lakes and reservoirs across the contiguous United States (US). Sampling took place in over 800 lakes and reservoirs from 1972 to 1975, and included a variety of physical, chemical, and biological metrics including data on nutrients and nutrient loading, hydrologic retention time, morphometry, and plankton community diversity.  Unlike current EPA National Lake Assessments that select a random sample of lakes across the US, the NES  targeted only lakes impacted directly or indirectly by municipal sewage treatment plant discharge \citep{nla-methods, nes-methods}. Until recently, these data were only available in their entirety as four separate scanned reports representing the northeastern and northcentral  (northeastern), eastern and southeastern (southeastern), central, and western regions of the US (Figure 1). In the remainder of the present paper we refer to the former two regions as simply the northeastern and southeastern regions.

@@

To our knowledge, there have been no attempts to transcribe the data into a usable, searchable digital database despite its use in previous studies.   For example, large portions of the dataset were used to examine large scale relationships between residence time and phytoplankton abundance \citep{soballe1987large}. Also, it was used to predict eutrophication incidence in a Bayesian framework (Lamon and Stow 2004). Smaller portions of the data were used to explore drivers of nutrient loading \citep{stomp2011large,Brettreviewreassessmentlake2007}. Yet, to our knowledge, the only study to use the NES dataset and provide a publicly available data supplement is \citet{stomp2011large}, and their data supplement was limited to a small subset of the available variables relating to phytoplankton community diversity.

@@

Data was collected from multiple locations within the water column and included in-situ measurements as well as laboratory analysis. Flow estimates and drainage area calculations were provided by the USGS and were determined from flow gages when present.  More detailed information on sampling methods, units, equipment, and accuracy can be found in the EPA survey methods publication \citep{nes-methods}. Due to historical nature of the dataset, the NES sampling design differs from more modern efforts \citep{nla-methods}. For example, the original NES data was collected from four separate regions of the US over the course of four years, whereas current assessments complete nation-wide sampling in a single summer. As such, NES data values represent the median of measurements taken in the spring, summer, and fall in either 1972 (northeastern), 1973 (southeastern), 1974 (central), or 1975 (western) rather than measurements taken in a single year.

We obtained the NES archival scanned reports from the EPA National Service Center for Environmental Publications (available at: https://www.epa.gov/nscep). The data for each NES region is contained in four separate files. We extracted the data from each file using automated techniques followed by manual quality assurance and checking. To begin, we enhanced (de-noised) each file using the local adaptive filtering algorithm as provided by the Imagemagick program (v6.8.9-9, available at: https://www.imagemagick.org/). Next, we processed the enhanced files using the Tesseract optical character recognition program (OCR) \citep{R-tesseract, smith2007overview}. The output of these initial extraction steps were recorded in a set of “raw data” files where each file contains the raw unprocessed text of each document page. The contents of specific fields in the raw data were extracted to a database using the automated rules provided by the nesR software package \citep{R-nesR}. Finally, all values in the database were manually checked for accuracy against the original scanned reports. Inaccurate OCR outputs were hand-corrected in the final database.

We provide the final dataset in an open non-proprietary format (comma-delimited, *.csv).  We generated metadata descriptions from the contents of the original scanned reports. All calculations, table construction, and figure generation were performed in R and saved as reproducible R scripts \citep{R-core}. Table and figure generation was accomplished with the use of the reshape2, plyr, and sp packages \citep{R-plyr,R-sp}.

@@

We have demonstrated an approach for rescuing historical data from scanned documents. In particular, our approach involved a two-step process of automated data scraping followed by hand-curation and quality assurance. Overall, we found that optical character recognition was an efficient method for reducing the labor associated with transcribing analog text records \citep{drinkwater2014use}. Unfortunately, optical character recognition technology does not have absolute accuracy. In our case, transcription was hampered by poor print and scan quality of the source paper documents. We discovered through our manual validation procedure that the OCR computations produced inaccurate values in approximately 6 - 17\% of the cells in the complete dataset (n = 4836). We expect that accuracy could be improved by experimenting with varying the window size of the local adaptive thresholding algorithm relative to the document font size. Our ability to experiment with thresholding window size was limited due to the computationally expensive nature of these extractions.

The end result of our approach was data from every lake and nearly every variable in the NES survey dataset. The only primary subset of the NES data which is not included in our final product is the phytoplankton distribution data which has already been digitally transcribed by \citet{stomp2011large}. The results of the present study could be used to explore anthropogenic and environmental drivers of lake eutrophication as well as to verify previously documented trends. One example is the 2007 National Lakes Assessment (NLA) Report, which included a reanalysis of some of the NES study lakes \citep{nla-methods}. This reanalysis considered population level trends in the NES lakes but did not consider trends in individual lakes or potential environmental drivers contributing to observed trends. On a population basis, the NLA reanalysis found that less than 30\% of the NES lakes had increased chlorophyll and phosphorus concentrations. The results of the present study could be used to verify these claims as well as to compare the NES data with more recent work such as the 2012 National Lakes Assessment.  Note that sampling techniques may differ from current techniques, so care should be given when making comparisons. In addition to their utility in validating historical trends, this dataset has value because it contains data on a number of hydrographic variables which are difficult to estimate such as water residence (retention) time. Such data is critical to a variety of hydrological and water quality modelling efforts \citep{Brettreviewreassessmentlake2007}.

Although our goal was to digitally transcribe the full NES dataset to facilitate studies on historical nutrient loading, it is worth noting the similarities between the present study and other scientific record digitization initiatives. Such initiatives are common in the climate and ocean sciences but they are just starting to gain momentum in the biological sciences \citep{allan2011international, freeman2017icoads}. To our knowledge, the present study is the first large-scale attempt at digitization of historical limnology records. We hope that by making our analysis open and reproducible we will inspire future efforts at recovering important records from the pre-digital era.

@@

\codedataavailability{
Original scanned reports from the EPA are available from the EPA National Service Center for Environmental Publications (https://www.epa.gov/nscep).  Our cleaned and useable data are available for download at doi:10.5063/F10G3H3Z. The data are provided as a zip file which contains all versions of the data including the raw and quality checked versions \citep{nes}. Moreover, the R package and R code used to scrape and analyze the data are provided by \citet{R-nesR} so that the methods may be reproduced and openly available for (re)use. All figures and summary statistics were generated by R scripts available in the data supplement linked above.
} %% use this section when having data sets and software code available
